Red Wine Exploratory Data Analysis by Niall O'Riordan
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(GGally)
library(RColorBrewer)
library(gridExtra)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
wines <- read.csv('wineQualityReds.csv')
```

#Introduction.

The data set explored contains tidy data about red wine samples.  It contains 
1599 observations of 11 input variables (chemical make-up based on physico-
chemical tests) and one output variable (quality).  The output is based on 
sensory data; the median of at least 3 evaluations made by wine experts on a 
scale of 0 (bad) to 10 (good).  More information can be found in the file 
descriptions.txt.

#Initial Analysis.

The first thing I did was run ggpairs() on the dataset (excluding X and q.cat),
in order to get the correlation figures for all the pairs.  This can be 
referenced below in the Bivariate Analysis section.

The chemicals with the highest correlation are alcohol, volatile acids, 
sulphates and citric acids respectively.  None are especially high with alcohol 
approaching a moderate level of correlation at .4796 and volatile acids 
meaning full but small at -0.391.  It's likely I'll also investigate sulphates
at .251 and citric acids at .226.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Exploration_1}
vvqplot <- function(varname, binwidth = NULL) {
  return(qplot(x = wines[varname], binwidth = binwidth))
} #using this function causes grid.arrange error - investigate later
#possibly add xlab()

up1 <- qplot(x = alcohol, data = wines) 
up2 <- qplot(x = volatile.acidity, data = wines) 
up3 <- qplot(x = sulphates, data = wines) 
up4 <- qplot(x = density, data = wines) 
  
grid.arrange(up1, up2, up3, up4, ncol = 1) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Exploration_2}
#vvqplot(wines$citric.acid) #still causing errors

qplot(x = citric.acid, data = wines)

```

I don't think there's many insights to be gained from the univariate plots.  
Most seem to have very small ranges and are very left sided.  Alcohol seems to
have a slightly broader range and I'll be investigating this relationship later.

Perhaps one interesting plot is the citric acid plot above.  There seems to be 
a large amount of wines with low levels of citric acid.  This might explain the
low correlation number; it's not so much that quality wines have a low level
of citric acid, as almost all the wines have a low level of citric acid.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Exploration_3}
ggplot(wines, aes(x = factor(quality))) +
   geom_bar() +
  labs(x="Quality Score", y="Count")
```

Before moving on to the next section, I'm going to add a "q.cat" variable to 
the dataset, with values "good" for wines rated 7-10, "fair" for "5-6", and 
"poor" for the rest.  I think most users won't care about the specific rating
of a wine, only whether it's a good wine or not.

I had planned on using the values 8-10 for "good", but looking at the histogram
above, it seems there's few wines in this category.  This raises some questions
about the dataset.  The vast majority of wines were in the 5-6 range.  This can
possibly be explained by the similarity of the wines being sampled.  I also 
wonder how the wines were rated, and if there was an outside influence on the 
judges?  For example, for the purposes of this project I saw no issue with 
categorising some wines as "poor", but if the wine producers were present 
during the rating, the judges may have been reluctant to rate the wine poorly.

I considered removing the "X" variable as it seems pretty redundant.  I decided
to keep it in case I sort the dataframe later and want to return it to it's
original order.

```{r echo=FALSE, message=FALSE, warning=FALSE, Add_Category_variable}
wines$q.cat <- cut(wines$quality, breaks = c(0, 4, 6, 10), 
                   labels=  c("poor", "fair", "good")) 
```

As can be seen below, categorising the wines makes the lack of variety in the 
quality ratings even more apparent.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Exploration_4}
qplot(x = q.cat, data = wines) 
```

#Bivariate Analysis

I'll start by running ggpairs() to try and get a feel for the relationships 
between the variables.  I've included the pdf ggpairsplot.pdf because the 
output below may be difficult to see.  The univariate plots appear on the 
diagonal, with bivariate plots below.


```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_G1}
#no need to have index column or q.cat
ggpairs(subset(wines, select = -c(X, q.cat)))
```

I didn't expect this to be very useful, and in my opinion it has turned out not
to be.  We're only interested in what chemicals affect the quality rating of
the wine, so graphing the rest of the relationships is pretty much a waste of 
time.  The plots on the horizontal 'quality line' are also don't' seem very 
useful.  All have discrete horizontal lines due to the 1-10 discrete quality 
scale, and all seem to have relatively wide distributions.  All also seem to 
suffer from over plotting.  I may try to reduce that to get more useful graphs 
later.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_A1}
ggplot(aes(x = alcohol, y = quality), data = wines) +
  xlim(quantile(wines$alcohol, 0.05), quantile(wines$alcohol, 0.95)) +
  geom_jitter(aes(color = quality))
```

Above scatter plot doesn't seem very useful.  Comparing the light blue to the 
dark, it is possible to see more light blue (high quality) observations at 
higher alcohol levels, but maybe I should do a box plot instead.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_A2}
ggplot(aes(x = quality, y = alcohol, group = q.cat), data = wines) +
  geom_boxplot(aes(color=q.cat))
```

The box plot above makes the correlation between quality and alcohol much easier
to see.  To the judges linking quality and alcohol could be subconscious, though
as we can see from some of the outliers in the "fair" it's not universal.  I've
read that some scientists theorise that our brains respond in a similar way to 
alcoholic sugars as regular sugar, so I kind of expected this.

I am surprised how close the "fair" and "poor" categories are, and expected a
more linear progression.  Some summary statistics are below, but don't seem to 
reveal much.  I'd hoped to explore this further, wondering if a combination of
chemicals makes the difference between "good" wines with high alcohol content 
and "fair" wines with high alcohol content, but didn't spend the time on it.

Summary of Alcohol Content (poor wines).

```{r echo=FALSE, message=FALSE, warning=FALSE }
summary(subset(wines$alcohol, wines$q.cat == "poor"))
```

Summary of Alcohol Content (fair wines).

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(wines$alcohol, wines$q.cat == "fair"))
```

Summary of Alcohol Content (good wines).

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(wines$alcohol, wines$q.cat == "good"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_V1}
ggplot(aes(x = volatile.acidity, y = quality), data = wines) +
  xlim(quantile(wines$volatile.acidity, 0.05), quantile(wines$volatile.acidity, 0.95)) +
  geom_point(aes(color = quality))
```

Again, the above scatter plot doesn't seem very useful.  It's possible to see 
a lot more light blue observations at lower volatile acidity levels than dark
blue, but I think I'll do a box plot again.


```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_V2}
ggplot(aes(x = quality, y = volatile.acidity, group = q.cat), data = wines) +
  geom_boxplot(aes(color=q.cat))

```

As can be seen in the box plot above, there appears to be a negative 
correlation between fixed acidity and quality; as volatile acidity decreases, 
quality increases.  This is to be expected, as wines with high levels of 
volatile acidity are described as tasting "vinegary".  Some summary statistics
are below but don't seem to reveal much.  It may be interesting to explore
some of the "good" quality wines with high volatile acidity levels to see if
we can identify if other chemicals are masking the taste.

Summary of Volatile Acidity (poor wines).

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(wines$volatile.acidity, wines$q.cat == "poor"))
```

Summary of Volatile Acidity (fair wines).

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(wines$volatile.acidity, wines$q.cat == "fair"))
```

Summary of Volatile Acidity (good wines).

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(wines$volatile.acidity, wines$q.cat == "good"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_S1}
ggplot(aes(x = sulphates, y = quality), data = wines) +
  xlim(quantile(wines$sulphates, 0.05), quantile(wines$sulphates, 0.95)) +
  geom_point(aes(color = quality), alpha = 1/4)
```

The above scatter plot shows the relationship between sulphates and quality.
I've attempted to make this plot more readable by introducing an alpha level of
1/4.  This helps, but the vast amount of average observations still present 
make the effect of quality increasing as sulphates do hard to see.  

In the box plot below I think the relationship is again easier to see.  This 
time I've also taken off the top and bottom 5% of sulphate observations to 
make the interquartile range appear bigger.  It's worth noting though how small
the variance in levels is.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_S2}
ggplot(aes(x = quality, y = sulphates, group = q.cat), data = subset(wines, 
            sulphates >= quantile(wines$sulphates, 0.05) &
            sulphates <= quantile(wines$sulphates, 0.95)) ) +
  geom_boxplot(aes(color=q.cat))
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_C1}
ggplot(aes(x = citric.acid, y = quality), data = wines) +
  xlim(quantile(wines$citric.acid, 0.05), quantile(wines$citric.acid, 0.95)) +
  geom_point(aes(color = quality), alpha = 1/4)
```

The above scatter plots show the relationship between citric acids and quality.
Like the previous plots I've attempted to make this plot more readable by 
introducing an alpha level of 1/4.  This again makes the higher number of light
blue observations at higher citric acid easier to see, but is still somewhat 
masked by the large amount of average quality observations.

In the box plot below I think the relationship is again easier to see.  This 
time I've again taken off the top and bottom 5% of citric acid observations to 
make the interquartile range appear bigger.  It's worth noting again how small
the variance in levels is.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_C2}
ggplot(aes(x = quality, y = citric.acid, group = q.cat), data = subset(wines, 
            citric.acid >= quantile(wines$citric.acid, 0.05) &
            citric.acid <= quantile(wines$citric.acid, 0.95)) ) +
  geom_boxplot(aes(color=q.cat))

```


```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_M1}
ggplot(aes(x= pH , y = fixed.acidity), data = wines) +
  geom_point() +
  geom_smooth(method = 'lm', color = 'red')

```

I didn't really find much interesting in the relationships between the other 
variables.  It may be useful in looking for erroneous data that some of the 
relationships should be linear.  In the example above, the strongest relation-
ship I found with a negative correlation of ???0.683 was PH v fixed acidity. This
is expected; as acidity levels go down, PH goes up.  The variation in the 
relationship can be explained by the other chemicals - the amount of chlorides,
etc.

I was surprised by the residual sugar content having little effect on the 
quality of the wine.  Like the alcohol content, I expected the judges to have
a subconscious reaction to the sugar.  It appears from the graph below that the 
residual sugar levels in almost all the wines are pretty low, and drop off for
the very high quality wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_M2}
ggplot(aes(x = quality, y = residual.sugar), data = wines) +
  geom_point(alpha = 1/4)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_M3}
ggplot(aes(x = alcohol, y = density), data = wines) +
  geom_point() +
  geom_smooth(method = 'lm', color = 'red')
```

Another relationship I found interesting was between density and alcohol with 
an actual value of ???0.496.  I would have expected the opposite and I think this
reflects my poor knowledge of how wines are produced.  I'd have expected the 
wine to get denser as it ferments, but perhaps it is decanted during the 
process.

# Multivariate Analysis

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_1}

ggplot(aes(x = quality, y = alcohol, color = volatile.acidity), data = wines) +
    geom_jitter(alpha = 1/3, size = 3) +
    scale_colour_gradient2(low = "red", high = "blue",
                           mid="#9933CC",midpoint = 0.8) +
    geom_smooth(method = 'lm', color = 'yellow')
```

The above plot is a bit noisy, but I think shows the main relationships that
affect wine quality.  The yellow line shows a general upward trend for quality
increasing as alcohol increases.  We can also see a lot more blue (high vol-
atile acidity at the left hand side of the plot, indicating acidity decreases
as quality increases.  I've added jitter to this plot and the next few to try
and make the scale look a bit more continuous.

We can see similar effects with sulphates below.  Sulphates help prevent a wine
oxidising, so as expected we see quality increase as sulphates do.  This affect
is limited though as too much sulphates results in a wine that is too dense.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_2}

ggplot(aes(x = quality, y = sulphates, color = density), data = wines) +
geom_jitter(alpha = 1/3, size = 3) +
scale_colour_gradient2(low = "red", high = "blue",
                       mid="#9933CC",midpoint = 0.997) +
geom_smooth(method = 'lm', color = 'yellow')

```

Similar effects can also be seen with citric acids below.  They are said to add
"freshness" to wine, but are illegal to add to wine in the EU except in 
specific circumstances.  I had hoped to do more analysis of this, but the 
citric acid levels are so low, I don't think it's worth continuing.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_3}
ggplot(aes(x = quality, y = citric.acid, color = pH), data = wines) +
    geom_jitter(alpha = 1/3, size = 3) + 
    scale_colour_gradient2(low = "red", high = "blue",
                           mid="#9933CC", midpoint = 3.31) +
    geom_smooth(method = 'lm', color = 'yellow')
```


# Final Plots and Summary

I think these are the three most useful plots that I found :

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(wines, aes(x=factor(quality), fill=I("blue"), col=I("black"))) + 
  geom_bar() +
  labs(title="Quality Score Counts") +
  labs(x="Quality Score", y="Count")
```

### Description One

The first plot is a simple histogram of the quality ratings.  This shows some 
of the limitations of the data set.  The ratings for quality are discrete, 
rather than continuous.  There's also a disparate number of 'average' ratings 
in the 5-6 range.  The similarity of quality observations might be in part 
explained by the fact that all the observations are of the Portuguese 
"Vinho Verde" red wine.

It's worth noting also that there are only six different scores observed which 
I've divided into three quality categories.  Wines that scored 3 or 4 are in 
the "poor" category, wines that scored 5 or 6 are in the "fair" category, and
wines that scored 7 or 8 are in the "good" category.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(x = quality, y = alcohol, group = q.cat), 
       data = subset(wines, alcohol >= quantile(wines$alcohol, 0.05) &
                        alcohol <= quantile(wines$alcohol, 0.95)) ) +
  geom_boxplot(aes(color=q.cat)) +
  labs(title="Alcohol Levels") +
  labs(x="Quality Score", y="% Alcohol")

```

### Description Two

This is similar to the earlier box plot of alcohol vs quality.  I've cut off
the top and bottom 5% of outliers to better show the relationship.  This makes
it easier to see the interquartile alcohol range of the "good" quality wines 
is higher than "fair"" or "poor".

Alcohol has a correlation of .4796 with quality which is moderate; as alcohol
levels go up, quality scores generally go up too.  This effect seems largest
with the "good"" quality wines.  The mean alcohol percentage of "poor" and 
"fair" wines are 10.22 and 10.25 respectively, but the mean for "good" wines
is 11.52.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x = quality, y = volatile.acidity, group = q.cat), 
       data = subset(wines, 
            volatile.acidity >= quantile(wines$volatile.acidity, 0.05) &
            volatile.acidity <= quantile(wines$volatile.acidity, 0.95)) ) +
  geom_boxplot(aes(color=q.cat)) +
  labs(title="Volatile Acidity Levels") +
  labs(x="Quality Score", y="Volatile Acidity g / dm^3")

```

### Description Three

This is again similar to the earlier box plot of volatile acidity vs quality. 
I've cut off the top and bottom 5% of outliers to better show the relationship.
This makes it easier to see the interquartile volatile acidity range of the 
"good"" quality wines is lower than "fair"" or "poor".

Volatile acidity has a negative correlation of -0.391 with quality which is 
small but meaningful; as volatile acidity levels go down, quality scores 
generally go up.  This effect seems more linear than what we see with alcohol.
The mean volatile acidity level of "poor" wines is .72, the mean for "fair" 
wines is .54, and the mean for "good" wines is .4.

------

# Reflection

The dataset is relatively small with relatively few inputs, so it was easy to
identify the main variables that affect the quality rating.  The largest was 
alcohol with a moderate correlation.  The next largest was volatile acidity
with a small but meaningful correlation.  

The dataset itself isn't very diverse.  The majority observations are at average
quality levels and the quality scores themselves are discrete integers.  I've 
used jitter in some of the plots to try and make the relationships seem more 
linear.  I also added a quality category (q.cat) variable to separate quality 
scores into ranges.  Not only do I think most people would be more interested 
in what category a wine is in rather than a numeric score, but I feel this made
it easier to see the relationships between the variables.

This project is already overdue, but I wish I had more time to explore the 
data.  I'd have liked to explore the main inputs (alcohol and volatile acidity)
that affected quality more.  I believe I would have needed to research the 
other inputs themselves more, to see if I could find information on how they
typically affect each other.  Some relationships are obvious (e.g. : residual 
sugar and density), but the relationship between sulphates, SO2 and acidity 
levels could be more complex, and combine to affect the quality rating.

In the future I'd like to explore larger data sets with more diverse wines. 
Though I'm unsure if I'd find the same insights, I feel this dataset was very 
limited in it's scope, and if I had developed a prediction model from it, I 
would worry that it would only be applicable to the type of wine studied.  I
think it would also be more useful to introduce a more continuous quality
rating scale.  The ratings are described as an average of ratings by 3 or 
more experts; there must have been some detail lost in the process.
